\vspace{2cm} \LARGE \bf Seasonal influenza circulation patterns and projections for 2017-2018

### 2017-02-24 auspice/reports/tex/feb-2017.tex
Update report text

% text layout
\usepackage{geometry}
\geometry{textwidth=16.75cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

@@

\date{February 16, 2017}

@@

Of the clades that emerged from the Texas/2012 background in early 2014, only 3c2.a currently remains at globally high prevalence (Fig.\ \ref{H3N2_clades}).
3c3.b viruses have scarcly been seen since the end of 2015.
While the uptick of 3c3.a viruses in mid-2016 in the US has failed to carry forward to the 2016-2017 US winter season.
Oceania still harbors 3c3.a viruses, but we fully expect these to be replaced by another clade in the coming year as Oceania is reseeded from the global reservoir.

@@

Within clade 3c2.a, a number of subclades with signature mutations at N171K, T131K/R142K, R142K, N121K/R144K, Q197K have emerged (Fig.\ \ref{H3N2_tree}).
Clades with R142K, N121K/R144K, Q197K arose in late 2014 or early 2015 but have not come to dominate.
However, the 171K clade that was first observed in mid-2015 has continued to increase in frequency at a moderate rate and is now at a global frequency of approximately 60\%.
Within the 171K clade, a subclade bearing the 121K mutation emerged in at the beginning of 2016 and has steadily increased in frequency now making of the majority of the parent 171K clade.

@@

Currently, the only noteworthy sister clade of 171K is the clade with mutations T131K/R142K (Fig.\ \ref{H3N2_mutations}).
This clade was first observed mid-2016 in China and dominated in China during last 3 month according to sequences available in GISAID, making up ~80\% of Chinese isolate since October.
It has been rising rapidly and has spread to Europe and North America.
In the branch leading up to this clade, we observe a rapid succession of mutation in codon 131 from threonine to asparagine and then to lysine and the mutation R142K.
Rapid succession of multiple mutations in the same codon suggest that these changes are adaptive.
Interestingly, we observe a large number of mutations to positively charged residues and particularly many parallel mutations to lysine at positions 121, 131, 142, 144, 171, and 197.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/h3n2_frequencies.pdf}
	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{H3N2_mutations}
\end{figure}

@@

@@

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/h1n1pdm_frequencies.pdf}
	\caption{\textbf{Frequency trajectories of 6b.1 subclades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{H1N1pdm_mutations}
\end{figure}

@@

We observe the continued dominance of the clade bearing mutations N129D, V146I and I117V (Fig.\ \ref{Vic_mutations}).
There is very little HA amino acid variation within the 117V viruses with no subclades rising to appreciable frequency.

@@

@@

>>>>

% text layout
\usepackage{geometry}
\geometry{textwidth=17cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

@@

\date{February 23, 2017}

@@

Of the clades that emerged from the Texas/2012 background in early 2014, only 3c2.a currently remains at globally high prevalence (Fig.\ \ref{H3N2_clades}).
3c3.b viruses have scarcly been seen since the end of 2015.
The uptick of 3c3.a viruses in mid-2016 in the US has failed to carry forward to the 2016-2017 US winter season.
Oceania still harbors 3c3.a viruses, but we fully expect these to be replaced by another clade in the coming year as Oceania is reseeded from the global reservoir.

@@

Within clade 3c2.a, a number of subclades have emerged (Fig.\ \ref{H3N2_tree}).
Notable clades include N171K (clade 3c2a.1), N171K/N121K, N121K/S144K, 131K/142K and R142K/Q197K.
Clades with R142K, N121K/R144K, Q197K arose in late 2014 or early 2015 but have not come to dominate.
However, the 171K clade (3c2a.1) that was first observed in mid-2015 has continued to increase in frequency at a moderate rate and is now at a global frequency of approximately 60\%.
Within the 171K clade, a subclade bearing the 121K mutation emerged in at the beginning of 2016 and has steadily increased in frequency now making of the majority of the parent 171K clade.

@@

Currently, the only noteworthy sister clade of 171K is the clade with mutations T131K/R142K (Fig.\ \ref{H3N2_mutations}).
This clade was first observed mid-2016 in China and dominated in China during last 3 month according to sequences available in GISAID, making up ~80\% of Chinese isolate since October.
It has been rising rapidly and has spread to Europe and North America.
In the branch leading up to this clade, we observe a rapid succession of mutation in codon 131 from threonine to asparagine and then to lysine and the mutation R142K.
Rapid succession of multiple mutations in the same codon suggest that these changes are adaptive.
Interestingly, we observe a large number of mutations to positively charged residues and particularly many parallel mutations to lysine at positions 121, 131, 142, 144, 171, and 197.
Clades N171K/N121K and T131K/R142K are on opposing genetic backgrounds so that there is competition between these clades; both cannot succeed.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{../figures/feb-2017/h3n2_frequencies.pdf}
	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{H3N2_mutations}
\end{figure}

@@

 \textit{Given current patterns of clade growth and decline, we predict clades 171K/121K and 131K/142K to be the most successful of currently circulating clades. We expect both to increase in the coming months. Longer term projections are difficult. 171K/121K may continue to dominate, 131K/142K may displace 171K/121K or another mutation may appear that determines the eventual outcome.}

@@

\textit{Given the lack of competition, we predict the continued dominance of 6b.1 viruses.}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/h1n1pdm_frequencies.pdf}
	\caption{\textbf{Frequency trajectories of 6b.1 subclades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{H1N1pdm_mutations}
\end{figure}

@@

We observe the continued dominance of the clade bearing mutations N129D, V146I and I117V (Fig.\ \ref{Vic_mutations}).
There is very little HA amino acid variation within the 117V viruses with no subclades rising to appreciable frequency.
There is deletion at AA site 162 that first appeared in mid-2016, but remains at low global frequency.

@@

\textit{Given the lack of competition, we predict the continued dominance of 117V viruses.}

@@

\textit{Genetic variation within 172Q/251V viruses is beginning to develop, but given lack of credible competition, we expect 172Q/251V to continual to dominate in the coming months.}
